---
title: "`r params$report_title`"
author: "`r params$report_author`"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
params:
  report_title: "Showcase repo functionality via plots"
  report_author: "Tamas Szabo"
knit: "(function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(dirname(inputFile)),'reports',paste(substr(basename(inputFile),1,nchar(basename(inputFile))-4),'.md',sep=''))) })"
---

# Setup

## Library imports

```{r include=FALSE, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)

library(ggplot2)
```

## Example dataset

A realistic dataset of colorectal and lung cancer mortality is used as an example input. The important difference compared to the actual data that can be downloaded is the 1-year granularity of age (instead of 5-year groups).

```{r message=FALSE, warning=FALSE}
example_dataset <- "../example_data/crc_lung_mortality.csv" %>%
  read.csv(check.names=FALSE) %>%
  pivot_longer(
    one_of("Colorectal (C18)", "Lung (C33-34)"),
    names_to = "Diagnosis", values_to = "N_cases"
  ) %>%
  mutate(
    Age = as.character(Age)
  )

# Commonly, numbers for both sexes merged are also present
example_dataset <- example_dataset %>%
  group_by(Age, Period, Diagnosis) %>%
  summarise(
    N_cases = sum(N_cases, na.rm=TRUE),
    Population = sum(Population, na.rm=TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    Sex = "Total"
  ) %>%
  bind_rows(example_dataset)

# Adding the total age group to show it does not interfere with models
example_dataset <- example_dataset %>%
  group_by(Sex, Period, Diagnosis) %>%
  summarise(
    N_cases = sum(N_cases, na.rm=TRUE),
    Population = sum(Population, na.rm=TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    Age = "Total"
  ) %>%
  select(Diagnosis, Age, Sex, Period, N_cases, Population) %>%
  bind_rows(example_dataset)

example_dataset %>%
  filter(Diagnosis == "Lung (C33-34)", Period == "2018") %>%
  filter(Sex == "Total", Age %in% seq(88, 90)) %>%
  head() %>%
  knitr::kable()
```

# Fundamental charts

## Standardized rates

The table output of the standardization step is expected to look as below:

```{r message=FALSE, warning=FALSE}
source("../scripts/standardized_rates.R", chdir=TRUE)

standardized_table <- example_dataset %>%
  calculate_standardized_rate(
    extra_grouping_vars="Diagnosis",
    standard_population = "2011"
  )

standardized_table %>%
  filter(Period == "2021", Sex != "Total") %>%
  select(Age, Sex, Diagnosis, value=Std_rate) %>%
  mutate(
    value = round(value, 2)
  ) %>%
  knitr::kable()
```

Graphical representation of the data would look something like:

```{r}
standardized_table %>%
  ggplot(aes(x=Period, y=Std_rate, color=Sex, group=Sex)) +
  geom_line() +
  geom_point() +
  facet_wrap("Diagnosis") +
  theme_bw() +
  theme(
    legend.position = "bottom"
  ) +
  labs(x="", y="Incidence rate projected to 2011", color="")
```

## Average annual change

```{r message=FALSE, warning=FALSE}
source("../scripts/annual_rate_change.R", chdir=TRUE)

change_table <- example_dataset %>%
  filter(Period %in% seq(2011, 2019)) %>%
  filter(Age != "Total") %>%
  execute_model_on_age_bins(
    seq(0, 90, 10), calculate_poisson_rate, grouping_vars=c("Sex", "Diagnosis")
  )

change_table %>%
  filter(Sex != "Total") %>%
  ggplot(aes(x=estimate, y=Age, color=Sex)) +
  geom_errorbarh(
    aes(xmin=CI_lo, xmax=CI_hi), position=position_dodge(width=0.6)
  ) +
  geom_point(position=position_dodge(width=0.6)) +
  facet_wrap("Diagnosis") +
  scale_x_continuous(labels=scales::percent) +
  theme_bw() +
  theme(
    legend.position = "bottom"
  ) +
  labs(x="", y="", color="")
```

## Impact of indidual years

As a first step, as well as a sanity check, calculate expected numbers first.

```{r message=FALSE, warning=FALSE}
source("../scripts/expected_case_numbers.R", chdir=TRUE)

expectation_table <- example_dataset %>%
  filter(Age != "Total", Sex != "Total") %>%
  execute_model_on_age_bins(
    seq(0, 90, 10), calculate_poisson_expectation, grouping_vars=c("Sex", "Diagnosis")
  )

expectation_table %>%
  filter(Age == "Total") %>%
  mutate(
    Period = as.numeric(Period)
  ) %>%
  ggplot(aes(x=Period)) +
  geom_col(aes(y=Expected_numbers), fill="grey") +
  geom_line(aes(y=N_cases, group=Sex), color="red") +
  facet_grid("Sex~Diagnosis")
```

As a next step, impact can be estimated as risk differences.

```{r message=FALSE, warning=FALSE}
source("../scripts/risk_difference.R", chdir=TRUE)

riskdiff_table <- expectation_table %>%
  filter(Period %in% seq(2020, 2021)) %>%
  pivot_longer(
    one_of("Expected_numbers", "N_cases"), values_to="N_cases"
  ) %>%
  calculate_risk_difference(
    c("Diagnosis", "Period", "Age", "Sex"),
    list(name=c("Expected_numbers", "N_cases"))
  )

riskdiff_table %>%
  filter(!Age %in% c("0-9", "10-19", "20-29", "30-39", "90-99")) %>%
  ggplot(aes(x=Riskdiff, y=Age, color=Sex)) +
  geom_errorbarh(
    aes(xmin=lower, xmax=upper), position=position_dodge(width=0.6)
  ) +
  geom_point(position=position_dodge(width=0.6)) +
  facet_grid("Period~Diagnosis") +
  scale_x_continuous(labels=scales::percent, limits=c(-1,1)) +
  theme_bw() +
  theme(
    legend.position = "bottom"
  ) +
  labs(x="", y="", color="")
```

An alternative would be to use the same model for extrapolation and assessment of the impact. The resulting change estimates should be similar to the ones above.

```{r message=FALSE, warning=FALSE}
source("../scripts/covid_year_impact.R", chdir=TRUE)

yearly_impact_table <- example_dataset %>%
  filter(Age != "Total", Period < 2022) %>%
  execute_model_on_age_bins(
    seq(0, 90, 10), calculate_impact_of_year,
    grouping_vars=c("Sex", "Diagnosis"), impacted_years = c(2020, 2021)
  )

yearly_impact_table %>%
  filter(Sex != "Total") %>%
  filter(!Age %in% c("0-9", "10-19", "20-29", "30-39", "90-99")) %>%
  filter(Modelparam %in% c("Period_2020", "Period_2021")) %>%
  mutate(
    Modelparam = gsub("Period_", "", Modelparam)
  ) %>%
  ggplot(aes(x=estimate, y=Age, color=Sex)) +
  geom_errorbarh(
    aes(xmin=CI_lo, xmax=CI_hi), position=position_dodge(width=0.6)
  ) +
  geom_point(position=position_dodge(width=0.6)) +
  facet_grid("Modelparam~Diagnosis") +
  scale_x_continuous(labels=scales::percent, limits=c(-1,1)) +
  theme_bw() +
  theme(
    legend.position = "bottom"
  ) +
  labs(x="", y="", color="")
```

# End